---
title: "BIOS735 Final Report"
author: "Group 5: Ji-Eun Park, Yueqi Shen, Jianqiao Wang, Yangjianchen Xu, Christina Zhou, Haotian Zou"
date: "4/27/2020"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# NOTES
Aim for Introduction to the dataset and the question, a Methods section with any important pre-processing etc., three sections of Results (which may also contain some method details), and Discussion about interpretation or possible future directions. 

# Introduction

# Dataset

We examined the `AIDS` dataset for our project, found in the JM package in R. This data is from a randomized clinical trial (RCT) examining two treatments in HIV patients. It contains both longitudinal and survival data on 467 patients who failed or were intolerant of zidovudine (AZT) therapy. These treatment groups of the RCT are two anti-retroviral drugs: zalcitabine and didanosine. The dataset contains a total of 1405 observations and 9 variables: patient identifier, time to death or censoring, death indicator, CD4 cell count, time points at which the CD4 cells count were recorded, treatment indicator, gender indicator, previous opportunistic infection (AIDS diagnosis) at study entry indicator, and AZT tolerance indicator.

# Aims

Our study aims are as follows:

(1). We will assess the association between CD4 cell count with time, adjusting for other covariates such as obstime, drug, gender, prevOI (indicator of previous opportunistic infection (AIDS diagnosis) at study entry), and AZT tolerance. We don't include the gender covariate, because more than 90% of subjects are Male, which leads to power loss when estimating gender coefficient.

(2). We will use the joint model to assess the association between survival outcome (death status and event time) with the drug efficacy and the latent mean of CD4 count at event time.

(3). We will assess and compare the model fitting result (mean and standard error) for the MCEM method and Bayesian method. We will also try Random Survival Forest. We will also do a simulation to compare the model fitting result.

(4). We will also propose a dynamic prediction framework to patients with some longitudinal outcomes available, for predicting the future longitudinal outcome and conditional survival probability at a given time.

# Methods 

Since this dataset contains both longitudinal and survival data, we proposed a joint model to simultaneously capture the random effects from the longitudinal data and to estimate the parameters of the survival model. A joint model is preferred to a two-step approach since previous studies have shown that the two-step approach may reduce efficacy which in turn inflates standard errors of estimates. For joint modelling, previous literature suggests that Bayesian methods are more efficient and computationally easier. We will compare frequentist methods to Bayesian methods by doing aim analysis with the Monte-Carlo EM (MCEM), Bayesian methods, and finally random survival forest (RSF) to compare to machine learning methods.

In the sections below, we describe in detail our model and methods.

### Model specification

Joint modelling framework:
$$
    Y_{ij} = m_i(t_{ij}) + \epsilon_{ij}
$$

$$
    h_i(t) = h_0(t)\exp(\boldsymbol{w_i'}\boldsymbol{\gamma} + \alpha m_i(t))
$$

where $m_i(t_{ij}) = \beta_0 + \beta_1 t_{ij} + \boldsymbol{x_{ij}}' \boldsymbol{\beta} + u_i$, and $m_i(T_i) = \beta_0 + \beta_1 T_i + \boldsymbol{x_i'^{(T_i)}} \boldsymbol{\beta^{(t)}} + u_i$.

We have $i=1,2, \dots, n$ subjects, $j = 1, 2,\dots, J_i$ visits for each subject. We observe the longitudinal outcome $Y_{ij}$ for subject i at visit j, and we observe the patient's covariate $\boldsymbol{x_{ij}}$ at time $t_{ij}$ . We also observe the event time $T_i = \text{min}(T_i^*, C_i)$, where $T_i^*$ is the actual failure time, and $C_i$ is the censoring time. We also observe the censoring indicator $\delta_i = I(T_i^*\leq C_i)$ , and covariates for survival model $\boldsymbol{w_i}$. 

Here, $m_i(t_{ij})$ is the latent longitudinal mean at time $t_{ij}$, and $u_i$ is the random effect for each subject $i$. And $h_i(t)$ is the hazard function for subject $i$ at time $t$, and $h_0$ is the baseline hazard function.

Assumption: 

$u_i \sim N(0, \sigma_u^2)$; 

$\epsilon_{ij} \sim N(0, \sigma_e^2)$; 

$Y_{ij} |u_i \perp Y_{ij'}|u_i$; 

$T_i^* \perp C_i$; 

$\{T_i, \delta_i \} | u_i \perp Y_{ij} | u_i$.

### Likelihood

We write the parameter space $\boldsymbol{\theta} = (\beta_0, \beta_1, \boldsymbol{\beta}, \sigma_u, \sigma_e, \boldsymbol{\gamma}, \alpha)'$. For simplicity of estimation, we posit a constant baseline hazard function $h_0$.

We write the full data likelihood:
$$
    L(\boldsymbol{\theta} | Y, T) = \prod_{i=1}^n \int_{-\infty}^{\infty} [\prod_{j=1}^{J_i} f(Y_{ij} | \beta, u_i)] [h_i(T_i)^{\delta_i} S_i(T_i)] f(u_i|\sigma_u^2) du 
$$

We write the complete data likelihood (assume random effect is known):
$$
    L_C(\boldsymbol{\theta} | Y, T) = \prod_{i=1}^n [\prod_{j=1}^{J_i} f(Y_{ij} | \beta, u_i)] [h_i(T_i)^{\delta_i} S_i(T_i)] f(u_i|\sigma_u^2) 
$$

We can write the complete data likelihood in Bayesian perspective:
$$
    L_B(\boldsymbol{\theta} | Y, T) = \prod_{i=1}^n [\prod_{j=1}^{J_i} f(Y_{ij} | \beta, u_i)] [h_i(T_i)^{\delta_i} S_i(T_i)] f(u_i|\sigma_u^2) P(\boldsymbol{\theta})
$$

where we denote $P(\boldsymbol{\theta})$ as the prior distribution of $\boldsymbol{\theta}$.

### Dynamic prediction

Suppose for a new subject $i'$, and we observe his longitudinal outcomes and covariates up to time $T$, where $j_{i'}$ is the number of visits up to time $T$ for subject $i'$. We want to predict his future longitudinal outcomes, and survival probability at time $T'$.

We can sample the random effect for the subject $i'$ based on its posterior distribution:
$$
    P(u_{i'} | Y_{i'}^{(T)}, T_{i'}>T, \boldsymbol{\hat \theta}) \propto \prod_{j=1}^{j_{i'}} f(Y_{i'j} |u_i', \boldsymbol{\hat \theta}) P(T_{i'}>T | u_{i'}, \boldsymbol{\hat \theta}) f(u_{i'} | \boldsymbol{\hat \theta})
$$

We sample from the above distribution, using Metropolis Hastings algorithm with random walk.

Suppose we have M samples of $u_{i'}$, we can then predict the future longitudinal outcome at time $T'$, and conditional survival probability at time $T'$, as:

$$
    m_{i'}(T') = \hat \beta_0 + \hat \beta_1 T' + \boldsymbol{x_{i'j}'} \boldsymbol{\hat \beta} + u_{i'}^{(d)} 
$$

$$
    P(T_{i'}>T' | T_{i'}>T) = S_{i'}(T' |  \boldsymbol{\hat \theta}, u_{i'}^{(d)})/S_{i'}(T  | \boldsymbol{\hat \theta}, u_{i'}^{(d)})
$$

We can calculate the area under curve (AUC) and Brier Score (BS) to compare the prediction accuracy. We will compare the model estimates, standard error, computation speed, and prediction accuracy. We will use cross-validation with 3:1 ratio to check prediction accuracy.

# Details in methods

We use MCEM algorithm, Bayesian methods, and Random Survival Forest to estimate parameters.

### MCEM algorithm

First, suppose we have the parameter estimate at iteration $t$. We sample the random effect $u_i$ from the posterior distribution of $u_i$, and do the maximization step.

We propose the Q-function as:

$$
    Q(Y|\theta^{(t)}) = E(l_c(\theta^{(t)}|Y, T)) = \sum_{i=1}^n \int l_i(\theta | Y_i, T_i) f(u_i | \theta^{(t)}, Y_i, T_i) du
$$

where $$l_i(\theta | Y_i, T_i) = \sum_{j=1}^{J_i} \log P(Y_{ij}|\theta, u_i) + \log[h_i(T_i)^{\delta_i}S_i(T_i)] + \log P(u_i |\theta)$$, 

$$f(u_i | \theta^{(t)}, Y_i, T_i) \propto \prod_{j=1}^{j_i} P(Y_{ij} |u_i, \boldsymbol{\theta^{(t)}}) h_i(T_i | \theta^{(t)})^{\delta_i}S_i(T_i | \theta^{(t)}) P(u_{i} | \boldsymbol{ \theta^{(t)}})$$.

We will use the Metropolis Hastings Algorithm with a random walk to draw the samples of random effect and calculate the integral by Monte Carlo method.

After we sample $M$ random effect for each subject, we get the augmented longitudinal dataset and survival dataset.

Second, we calculate $\sigma_u^{(t+1)} = \text{SD}(\boldsymbol{u})$, where $\boldsymbol{u}$ is a vector of random effects for all subjects. We calculate $\sigma_e^{(t+1)}$ by simply fitting a linear model using augmented longitudinal dataset, treating $\boldsymbol{u}$ as the offset.

Third, for maximization of $(\beta_0, \beta_1, \boldsymbol{\beta}, \log(h_0))$, we use MH algorithm with RW to sample from the log likelihood given $\sigma_u^{(t+1)}$, $\sigma_e^{(t+1)}$, $\gamma^{(t)}$, $\alpha^{(t)}$.

Finally, for maximization $\gamma, \alpha$, we fit a Cox proportional hazards model, using the augmented survival dataset, given $\log(h_0)^{(t+1)}$, $m_i(T_i)^{(t+1), m}$, where $m_i(T_i)^{(t+1), m} = \beta_0^{(t+1)} + \beta_1^{(t+1)}T_i + \boldsymbol{x_i'^{(T_i)}} \boldsymbol{\beta^{(t)}} + u_i^{m}$, and $u_i^{m}$ denotes the $m^{th}$ random effect for subject $i$.

### Bayesian methods

We will use Stan to perform sampling, where we will use No-U-Turn Sampler (NUTS), based on Hamiltonian Monte Carlo (HMC).

We will use normal prior distributions for $\beta$..., and inverse Gamma prior distributions for the variance parameters. We will use the following prior distribution for the parameters for Bayesian methods:

(1) $\beta \sim N(0, 10^2)$.

(2) $\log(h_0) \sim N(0, 10^2)$.

(3) $\gamma \sim N(0, 10^2)$.

(4) $\alpha \sim N(0, 10^2)$.

(5) $\sigma_u \sim \text{Inverse Gamma}(0.1, 0.1)$.

(6) $\sigma_e \sim \text{Inverse Gamma}(0.1, 0.1)$.

### Initial value specification

Good initial values will accelerate the convergence speed for both EM algorithm and Bayesian methods.

To get initial values of $\beta_0, \beta_1, \boldsymbol{\beta}, \sigma_u, \sigma_e$, we fit a linear mixed effects model, and extract the estimated coefficients and variance components.

To get initial values of $\log(h_0), \gamma, \alpha$, we fit an AFT model with exponential distribution, where the latent longitudinal mean is calculated at the event time for each subject. We extract the estimated intercept and coefficients, and take negative as initial values of $\log(h_0), \gamma, \alpha$.


### Random Survival Forest

For machine learning methods: We plan to use random survival forest, considering only for the survival model, and ignoring the longitudinal model, where we will use the CD4 count nearest to survival time as the substitute of latent longitudinal mean.

# Simulation

In order to ensure our methods are correct, we simulated data to test our package. Our simulation set-up is as follows:

First, we simulated the longitudinal responses using the following formula: $Y_{ij} = 20 + (-1)*t_{ij} + (-4)*x_i + u_i + \epsilon_{ij}$, where $u_i \sim N(0, 4), \epsilon_{ij} \sim N(0, 1), x_i \sim \text{binom}(1, 0.4)$, for subject $i$ at time point $j$ where $i = 1,\dots,N$ and $j = 0,2,6,12,18)$. 

Next, we simulated N random uniform distributions to correspond with the survival probabilities $S(T)$. We used this relation between $S(T_i)$ and the hazard to calculate the survival time. The failure time is defined as: $h_i(T_i) = \exp(-2) \exp(-1*w_i + (-0.2)*m_i(T_i))$, where $w_i$ corresponds to the drug variable in the `AIDS` dataset and $w_i \sim \text{binom}(1, 0.5), m_i(T_i) = 20 + (-1)*T_i + (-4)*x_i + u_i$. We calculated the survival time $T_i^*$ using following formula:

$$
S_i(t) = \exp(-\int_0^t h_i(t) dt) = \exp\{-h_0\exp[w_i \gamma + \alpha(\beta_0 + \beta_2x_i + u_i)] \frac{\exp(\alpha \beta_1 t) - 1}{(\alpha \beta_1)}
$$
Thus, we see that $T_i^* =\frac{1}{\alpha\beta_1}log[1-\frac{\alpha\beta_1\log(S_i(t))}{h_0\exp(\gamma w_i+\alpha(\beta_0+\beta_2x_i+u_i))}]$, where $\alpha = -0.2, \beta_0 = 20,\beta_1 = -1,\beta_2 = -4,\gamma = -1, h_0 = \exp(-2)$.

In order to obtain a death rate close to that of the `AIDS` dataset ($40.25\%$), we simulated the censoring indicator for each individual from a random Uniform distribution, $C_i \sim Unif(0,27)$. We define the time to event as $T_i = min(T_i^*, C_i)$. Finally, we define our death indicator as $\delta_i = I(T_i^* \le C_i)$. 

For each simulation, we produce two sets of joint data: one longitudinal dataset and one survival dataset. The survival dataset consists of one row per subject with variables time to event, death indicator, and treatment group (drug). The longitudinal dataset consists of multiple observations per subject with variables response, observed time, and covariates. Subjects do not have observation times beyond the event time.

For cross validation, we simulated N=500 subjects for the training dataset and another 200 subjects for the testing dataset.

# Results 

### Load our package `jm5` and separate AIDS dataset into a longitudinal dataset and a survival dataset

```{r}
library(knitr)
library(devtools)
load_all('jm5')
```

Define our longitudinal formula and survival formula, and separate datasets

```{r}
l.formula <- CD4 ~ obstime + drug + prevOI + AZT
s.formula <- Surv(Time, death) ~ drug
dat <- JM::aids
l <- jm_filter(l.formula, s.formula, dat, ID = dat$patient)
long.dat <- l[[1]]
surv.dat <- l[[2]]
```

Head of longitudinal dataset and survival dataset
```{r}
head(long.dat)
head(surv.dat)
```

### Summary statistics for AIDS dataset

```{r}

```

### Model fit for AIDS dataset - MCEM algorithm

```{r}
l.formula2 <- CD4 ~ obstime + drugddI + prevOIAIDS + AZTfailure
s.formula2 <- Surv(Time, death) ~ drugddI

mcem.result <- MCEM(l.formula2, s.formula2, long.dat, surv.dat, max.iter = 10)
mcem.result
```

### Model fit for AIDS dataset - Bayesian method

```{r}
fit.result <- fit_stan(l.formula2, s.formula2, long.dat, surv.dat)
summary.mean <- summary(fit.result)$summary[, 1]
summary.mean <- summary.mean[1:(length(summary.mean) - 1)]
summary.mean
```

### RSF for AIDS dataset

```{r}

```

### Cross-validation: AUC and BS for AIDS dataset (MCEM, Bayesian)

```{r}
cv.dat <- create_cv(long.dat, surv.dat)
st <- c(2, 4, 6) ## starting time
dt <- c(0.5, 1) ## prediction time window
mcem.auc.bs.mean <- matrix(0, length(st)*length(dt), 2)
stan.auc.bs.mean <- matrix(0, length(st)*length(dt), 2)
for (i in 1:length(cv.dat)){
  ## training data and testing data
  long.train <- cv.dat[[i]][[1]]
  long.test <- cv.dat[[i]][[2]]
  surv.train <- cv.dat[[i]][[3]]
  surv.test <- cv.dat[[i]][[4]]
  
  ## run MCEM and Stan on training dataset
  mcem.train.mean <- MCEM(l.formula2, s.formula2, long.train, surv.train, max.iter = 10, progress = FALSE)
  stan.train <- fit_stan(l.formula2, s.formula2, long.train, surv.train, progress = FALSE)
  stan.train.mean <- summary(stan.train)$summary[, 1]
  stan.train.mean <- stan.train.mean[-length(stan.train.mean)]
  
  ## Calculate AUC and BS ##
  mcem.auc.bs <- matrix(NA, length(st)*length(dt), 2)
  stan.auc.bs <- matrix(NA, length(st)*length(dt), 2)
  index <- 0
  for (Tstart in st){
    for (Tdelta in dt){
      index <- index + 1
      mcem.auc.bs[index, ] <- calc_AUC_BS(l.formula2, s.formula2, 
                                          long.test, surv.test,
                                          Tstart, Tstart+Tdelta, 
                                          mcem.train.mean)
      stan.auc.bs[index, ] <- calc_AUC_BS(l.formula2, s.formula2, 
                                          long.test, surv.test,
                                          Tstart, Tstart+Tdelta, 
                                          stan.train.mean)
    }
  }
  mcem.auc.bs.mean <- mcem.auc.bs.mean + mcem.auc.bs
  stan.auc.bs.mean <- stan.auc.bs.mean + stan.auc.bs
}

mcem.auc.bs.mean/length(cv.dat)
stan.auc.bs.mean/length(cv.dat)
```

### Cross-validation: AUC and BS for AIDS dataset (RSF)

```{r}

```

### Simulation: evaluate correctness of MCEM and Bayesian methods

```{r}

```
