---
title: "BIOS735 Final Report"
author: "Group 5"
date: "4/27/2020"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Results 

### Load our package `jm5` and separate AIDS dataset to longitudinal dataset and survival dataset

```{r include=FALSE}
library(knitr)
library(devtools)
library(tidyverse)
library(JM)
load_all('jm5')
```

Define our longitudinal formula and survival formula, and separate datasets

```{r}
l.formula <- CD4 ~ obstime + drug + prevOI + AZT
s.formula <- Surv(Time, death) ~ drug
dat <- JM::aids
l <- jm_filter(l.formula, s.formula, dat, ID = dat$patient)
long.dat <- l[[1]]
surv.dat <- l[[2]]
```

Head of longitudinal dataset and survival dataset
```{r}
head(long.dat)
head(surv.dat)
```

### Summary statistics for AIDS dataset

```{r echo=FALSE}
va = c("ID", "Time", "death", "CD4", "obstime", "drug", "prevOI", "AZT", "gender")
desc = c("Patients identifier (1-467)",
         "The time to death or censoring",
         "0 for censoring, 1 for death",
         "The CD4 cells count",
         "The time points at which the CD4 cells count was recorded (0, 2, 6, 12, 18)",
         "ddI (1) for didanosine, ddC (0) for zalcitabine",
         "AIDS (1) for previous opportunistic infection, noAIDS (0) for no previous infection",
         "failure (1) for AZT failure, intolerance (0) for AZT intolerance",
         "1 for male, 0 for female"
         )
summary_stat = data.frame(Variable = va, Description = desc)
kable(summary_stat)
```

```{r echo=FALSE, out.width='50%'}
summ = aids[,c(1,3:9)] %>%
  spread(key = obstime, value = CD4)
summ = summ[,2:6]
summ$drug = ifelse(summ$drug == "ddI", 1, 0)
summ$gender = ifelse(summ$gender == "male", 1, 0)
summ$prevOI = ifelse(summ$prevOI == "AIDS", 1, 0)
summ$AZT = ifelse(summ$AZT == "failure", 1, 0)
summ = gather(summ, death, drug, gender, prevOI, AZT,
              key = "variables", value = "count")
ggplot(summ, aes(variables)) +
  geom_bar(aes(fill = as.factor(count))) +
  theme_light() +
  labs(x = "Variable",
       y = "Count",
       title = "Summary plot for dichotomous variables") +
  scale_fill_discrete(name = NULL,
                      labels = c("intolerance/censoring/ddC/female/noAIDS",
                                 "failure/death/ddI/male/AIDS")) +
  theme(legend.position="top")

ggplot(surv.dat, aes(x = as.factor(ifelse(death == 0, "Censoring", "Death")), y = Time, fill = as.factor(drugddI))) +
  geom_boxplot() +
  labs(x = c("Status"),
       y = "Time to death or censoring",
       title = "Summary plot for survival part") +
  scale_fill_discrete(name = "drug",
                      labels = c("ddC","ddI")) +
  theme_light() +
  theme(legend.position="top")
```

```{r echo=FALSE, out.width='33.33%'}
ggplot(long.dat, aes(x = obstime, y = CD4, group = as.factor(ID), colour = as.factor(prevOIAIDS))) +
  geom_line() +
  scale_x_continuous(breaks = c(0,2,6,12,18)) +
  scale_color_discrete(name = "prevOI",
                       labels = c("AIDS", "noAIDS")) +
  labs(y = "CD4 cell counts",
       x = "Observation time",
       title = "Summary plot for longitudinal part (grouped by prevOI)") +
  theme_light() +
  theme(legend.justification=c(0.98,0.99), legend.position=c(0.98,0.99))
ggplot(long.dat, aes(x = obstime, y = CD4, group = as.factor(ID), colour = as.factor(drugddI))) +
  geom_line() +
  scale_x_continuous(breaks = c(0,2,6,12,18)) +
  scale_color_discrete(name = "drug",
                       labels = c("ddC", "ddI")) +
  labs(y = "CD4 cell counts",
       x = "Observation time",
       title = "Summary plot for longitudinal part (grouped by drug)") +
  theme_light() +
  theme(legend.justification=c(0.98,0.99), legend.position=c(0.98,0.99))
ggplot(long.dat, aes(x = obstime, y = CD4, group = as.factor(ID), colour = as.factor(AZTfailure))) +
  geom_line() +
  scale_x_continuous(breaks = c(0,2,6,12,18)) +
  scale_color_discrete(name = "AZT",
                       labels = c("intolerance", "failure")) +
  labs(y = "CD4 cell counts",
       x = "Observation time",
       title = "Summary plot for longitudinal part (grouped by AZT)") +
  theme_light() +
  theme(legend.justification=c(0.98,0.99), legend.position=c(0.98,0.99))
```

### Model fit for AIDS dataset - MCEM algorithm

```{r eval=FALSE, include=FALSE}
l.formula2 <- CD4 ~ obstime + drugddI + prevOIAIDS + AZTfailure
s.formula2 <- Surv(Time, death) ~ drugddI

mcem.result <- MCEM(l.formula2, s.formula2, long.dat, surv.dat, max.iter = 10)
mcem.result
```

### Simulation: evaluate correctness of MCEM and Bayesian methods

```{r include=FALSE}
## Simulation for Bayesian method
simulation.Bayesian = sim.Bayesian(sim.num = 10)
```
```{r echo=FALSE}
kable(simulation.Bayesian)
```

### Simulation: estimated AUC, true AUC and estimated BS (MCEM and Bayesian)

```{r}

```