---
title: "BIOS735 Final Report"
author: "Group 5"
date: "4/27/2020"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# NOTES
Aim for Introduction to the dataset and the question, a Methods section with any important pre-processing etc., three sections of Results (which may also contain some method details), and Discussion about interpretation or possible future directions. 

\

Introduction\

Project Aims\
Data description\
Methods\
-Processing the data\
-Model\
-3 things we plan to do (bayesian, MCEM, random survival forest)\
Results\
-3 things we plan to do (bayesian, MCEM, random survival forest)\
Discussion\

# Introduction

The dataset we will examine for our project can be found in the JM package in R. This data is from a randomized clinical trial (RCT) examining two treatments in HIV patients. It contains both longitudinal and survival data on 467 patients who failed or were intolerant of zidovudine (AZT) therapy. These treatment groups of the RCT are two anti-retroviral drugs: zalcitabine and didanosine. The dataset contains a total of 1405 observations and 9 variables: patient identifier, time to death or censoring, death indicator, CD4 cell count, time points at which the CD4 cells count were recorded, treatment indicator, gender indicator, previous opportunistic infection (AIDS diagnosis) at study entry indicator, and AZT tolerance indicator.

Since this dataset contains both longitudinal and survival data, we proposed a joint model to simultaneously capture the random effects from the longitudinal data and to estimate the parameters of the survival model. A joint model is preferred to a two-step approach since previous studies have shown that the two-step approach may reduce efficacy which in turn inflates standard errors of estimates. For joint modelling, previous literature suggests that Bayesian methods are more efficient and computationally easier.

# Aims

Our study aims are as follows:

(1). We will assess the association between CD4 cell count with time, adjusting for other covariates such as obstime, drug, gender, prevOI (indicator of previous opportunistic infection (AIDS diagnosis) at study entry), and AZT tolerance. We don't include the gender covariate, because more than 90% of subjects are Male, which leads to power loss when estimating gender coefficient.

(2). We will use the joint model to assess the association between survival outcome (death status and event time) with the drug efficacy and the latent mean of CD4 count at event time.

(3). We will assess and compare the model fitting result (mean and standard error) for the MCEM method and Bayesian method. We will also try Random Survival Forest. We will also do a simulation to compare the model fitting result.

(4). We will also propose a dynamic prediction framework to patients with some longitudinal outcomes available, for predicting the future longitudinal outcome and conditional survival probability at a given time.

# Methods 

### Model specification

Joint modelling framework:
$$
    Y_{ij} = m_i(t_{ij}) + \epsilon_{ij}
$$

$$
    h_i(t) = h_0(t)\exp(\boldsymbol{w_i'}\boldsymbol{\gamma} + \alpha m_i(t))
$$

where $m_i(t_{ij}) = \beta_0 + \beta_1 t_{ij} + \boldsymbol{x_{ij}}' \boldsymbol{\beta} + u_i$, and $m_i(T_i) = \beta_0 + \beta_1 T_i + \boldsymbol{x_i'^{(T_i)}} \boldsymbol{\beta^{(t)}} + u_i$.

We have $i=1,2, \dots, n$ subjects, $j = 1, 2,\dots, J_i$ visits for each subject. We observe the longitudinal outcome $Y_{ij}$ for subject i at visit j, and we observe the patient's covariate $\boldsymbol{x_{ij}}$ at time $t_{ij}$ . We also observe the event time $T_i = \text{min}(T_i^*, C_i)$, where $T_i^*$ is the actual failure time, and $C_i$ is the censoring time. We also observe the censoring indicator $\delta_i = I(T_i^*\leq C_i)$ , and covariates for survival model $\boldsymbol{w_i}$. 

Here, $m_i(t_{ij})$ is the latent longitudinal mean at time $t_{ij}$, and $u_i$ is the random effect for each subject $i$. And $h_i(t)$ is the hazard function for subject $i$ at time $t$, and $h_0$ is the baseline hazard function.

Assumption: 

$u_i \sim N(0, \sigma_u^2)$; 

$\epsilon_{ij} \sim N(0, \sigma_e^2)$; 

$Y_{ij} |u_i \perp Y_{ij'}|u_i$; 

$T_i^* \perp C_i$; 

$\{T_i, \delta_i \} | u_i \perp Y_{ij} | u_i$.

### Likelihood

We write the parameter space $\boldsymbol{\theta} = (\beta_0, \beta_1, \boldsymbol{\beta}, \sigma_u, \sigma_e, \boldsymbol{\gamma}, \alpha)'$. For simplicity of estimation, we posit a constant baseline hazard function $h_0$.

We write the full data likelihood:
$$
    L(\boldsymbol{\theta} | Y, T) = \prod_{i=1}^n \int_{-\infty}^{\infty} [\prod_{j=1}^{J_i} f(Y_{ij} | \beta, u_i)] [h_i(T_i)^{\delta_i} S_i(T_i)] f(u_i|\sigma_u^2) du 
$$

We write the complete data likelihood (assume random effect is known):
$$
    L_C(\boldsymbol{\theta} | Y, T) = \prod_{i=1}^n [\prod_{j=1}^{J_i} f(Y_{ij} | \beta, u_i)] [h_i(T_i)^{\delta_i} S_i(T_i)] f(u_i|\sigma_u^2) 
$$

We can write the complete data likelihood in Bayesian perspective:
$$
    L_B(\boldsymbol{\theta} | Y, T) = \prod_{i=1}^n [\prod_{j=1}^{J_i} f(Y_{ij} | \beta, u_i)] [h_i(T_i)^{\delta_i} S_i(T_i)] f(u_i|\sigma_u^2) P(\boldsymbol{\theta})
$$

where we denote $P(\boldsymbol{\theta})$ as the prior distribution of $\boldsymbol{\theta}$.

### Dynamic prediction

Suppose for a new subject $i'$, and we observe his longitudinal outcomes and covariates up to time $T$, where $j_{i'}$ is the number of visits up to time $T$ for subject $i'$. We want to predict his future longitudinal outcomes, and survival probability at time $T'$.

We can sample the random effect for the subject $i'$ based on its posterior distribution:
$$
    P(u_{i'} | Y_{i'}^{(T)}, T_{i'}>T, \boldsymbol{\hat \theta}) \propto \prod_{j=1}^{j_{i'}} f(Y_{i'j} |u_i', \boldsymbol{\hat \theta}) P(T_{i'}>T | u_{i'}, \boldsymbol{\hat \theta}) f(u_{i'} | \boldsymbol{\hat \theta})
$$

We sample from the above distribution, using Metropolis Hastings algorithm with random walk.

Suppose we have M samples of $u_{i'}$, we can then predict the future longitudinal outcome at time $T'$, and conditional survival probability at time $T'$, as:

$$
    m_{i'}(T') = \hat \beta_0 + \hat \beta_1 T' + \boldsymbol{x_{i'j}'} \boldsymbol{\hat \beta} + u_{i'}^{(d)} 
$$

$$
    P(T_{i'}>T' | T_{i'}>T) = S_{i'}(T' |  \boldsymbol{\hat \theta}, u_{i'}^{(d)})/S_{i'}(T  | \boldsymbol{\hat \theta}, u_{i'}^{(d)})
$$

We can calculate the area under curve (AUC) and Brier Score (BS) to compare the prediction accuracy. We will compare the model estimates, standard error, computation speed, and prediction accuracy. We will use cross-validation to check prediction accuracy.

# Details in methods

We use MCEM algorithm, Bayesian methods, and Random Survival Forest to estimate parameters.

### MCEM algorithm

First, suppose we have the parameter estimate at iteration $t$. We sample the random effect $u_i$ from the posterior distribution of $u_i$, and do the maximization step.

We propose the Q-function as:

$$
    Q(Y|\theta^{(t)}) = E(l_c(\theta^{(t)}|Y, T)) = \sum_{i=1}^n \int l_i(\theta | Y_i, T_i) f(u_i | \theta^{(t)}, Y_i, T_i) du
$$

where $$l_i(\theta | Y_i, T_i) = \sum_{j=1}^{J_i} \log P(Y_{ij}|\theta, u_i) + \log[h_i(T_i)^{\delta_i}S_i(T_i)] + \log P(u_i |\theta)$$, 

$$f(u_i | \theta^{(t)}, Y_i, T_i) \propto \prod_{j=1}^{j_i} P(Y_{ij} |u_i, \boldsymbol{\theta^{(t)}}) h_i(T_i | \theta^{(t)})^{\delta_i}S_i(T_i | \theta^{(t)}) P(u_{i} | \boldsymbol{ \theta^{(t)}})$$.

We will use the Metropolis Hastings Algorithm with a random walk to draw the samples of random effect and calculate the integral by Monte Carlo method.

After we sample $M$ random effect for each subject, we get the augmented longitudinal dataset and survival dataset.

Second, we calculate $\sigma_u^{(t+1)} = \text{SD}(\boldsymbol{u})$, where $\boldsymbol{u}$ is a vector of random effects for all subjects. We calculate $\sigma_e^{(t+1)}$ by simply fitting a linear model using augmented longitudinal dataset, treating $\boldsymbol{u}$ as the offset.

Third, for maximization of $(\beta_0, \beta_1, \boldsymbol{\beta}, \log(h_0))$, we use MH algorithm with RW to sample from the log likelihood given $\sigma_u^{(t+1)}$, $\sigma_e^{(t+1)}$, $\gamma^{(t)}$, $\alpha^{(t)}$.

Finally, for maximization $\gamma, \alpha$, we fit a Cox proportional hazards model, using the augmented survival dataset, given $\log(h_0)^{(t+1)}$, $m_i(T_i)^{(t+1), m}$, where $m_i(T_i)^{(t+1), m} = \beta_0^{(t+1)} + \beta_1^{(t+1)}T_i + \boldsymbol{x_i'^{(T_i)}} \boldsymbol{\beta^{(t)}} + u_i^{m}$, and $u_i^{m}$ denotes the $m^{th}$ random effect for subject $i$.

### Bayesian methods

We will use Stan to perform sampling, where we will use No-U-Turn Sampler (NUTS), based on Hamiltonian Monte Carlo (HMC).

We will use normal prior distributions for $\beta$..., and inverse Gamma prior distributions for the variance parameters. We will use the following prior distribution for the parameters for Bayesian methods:

(1) $\beta \sim N(0, 10^2)$.

(2) $\log(h_0) \sim N(0, 10^2)$.

(3) $\gamma \sim N(0, 10^2)$.

(4) $\alpha \sim N(0, 10^2)$.

(5) $\sigma_u \sim \text{Inverse Gamma}(0.1, 0.1)$.

(6) $\sigma_e \sim \text{Inverse Gamma}(0.1, 0.1)$.

### Initial value specification

Good initial values will accelerate the convergence speed for both EM algorithm and Bayesian methods.

To get initial values of $\beta_0, \beta_1, \boldsymbol{\beta}, \sigma_u, \sigma_e$, we fit a linear mixed effects model, and extract the estimated coefficients and variance components.

To get initial values of $\log(h_0), \gamma, \alpha$, we fit an AFT model with exponential distribution, where the latent longitudinal mean is calculated at the event time for each subject. We extract the estimated intercept and coefficients, and take negative as initial values of $\log(h_0), \gamma, \alpha$.


### RSF

For machine learning methods: We used the random survival forest (RSF), considering only for the survival model, and ignoring the longitudinal model, where we will use the CD4 count nearest to survival time as the substitute of latent longitudinal mean.

For the random survival forest, we use our function `RandomSurvivalForest`. This function `rfsrc` from the r package `randomForestSRC`, allows random forest for right-censored survival.

Here we use the log-rank test to grow survival trees. In our dataset, we split each tree at $x\le c$ and $x>c$ where x is CD4 and drug(ddI). Since we only use 2 variables, each split depends on one variable. Such a split defines left($x\le c$) and right($x>c$) daughter membership. Each split maximizes survival differences between the two nodes. Let $t_1<t_2<\dots<t_m$ be the distinct times of `death` in the parent node $h$, and for each left and right node, we get the number of deaths ($d_{k,l}$, $d_{k,r}$ where l=left, r=right) and patients at risk respectively at time $t_k$ ($Y_{k,l}$,$Y_{k,r}$.

$$
Y_{k,l} = \#\{i:T_i \ge t_k, x_i \le c\},\quad Y_{k,r} = \#\{i:T_i \ge t_k, x_i > c\}
$$

The log-rank test for each split at value $c$ for a varialbe $x$ is
$$
L(x,c) = \frac{\sum\limits_{k=1}^{m} (d_{k,l} - Y_{k,l}\frac{d_k}{Y_k})}{\sqrt{\sum\limits_{k=1}^m \frac{Y_{k,l}}{Y_k} (1-\frac{Y_{k,l}}{Y_k}) (\frac{Y_k-d_k}{Y_k-1}) d_k}}
$$

We find the c such that maximizes $|L(x,c)|$, which means thhe differences between the two nodes are maximized.


# Simulation

To ensure correctness of our methods, we use the simulation set-up as follows:

First, we simulate the longitudinal responses using the following formula: $Y_{ij} = 20 + (-1)*t_{ij} + (-4)*x_i + u_i + \epsilon_{ij}$, where $u_i \sim N(0, 4), \epsilon_{ij} \sim N(0, 1), x_i \sim \text{binom}(1, 0.4)$. 

Second, we simulate N random uniform distribution, which corresponds to the survival probability. And we calculate the failure time as: $h_i(T_i) = \exp(-2) \exp(-1*w_i + (-0.2)*m_i(T_i))$, where $w_i \sim \text{binom}(1, 0.5), m_i(T_i) = 20 + (-1)*T_i + (-4)*x_i + u_i$. And we can calculate the survival time $T_i^*$ using following formula:

$$
S_i(t) = \exp(-\int_0^t h_i(t) dt) = \exp\{-h_0\exp[w_i \gamma + \alpha(\beta_0 + \beta_2*x_i + u_i)] [\exp(\alpha \beta_1 t) - 1]/(\alpha \beta_1)\}
$$

We simulate N=500 subjects, and time points as (0, 2, 6, 12, 18) as similar as the `aids` data in the longitudinal model, and another 200 subjects as the testing dataset.

# Results 

### Load our package `jm5` and separate AIDS dataset to longitudinal dataset and survival dataset

```{r}
library(knitr)
library(devtools)
load_all('jm5')
```

Define our longitudinal formula and survival formula, and separate datasets

```{r}
l.formula <- CD4 ~ obstime + drug + prevOI + AZT
s.formula <- Surv(Time, death) ~ drug
dat <- JM::aids
l <- jm_filter(l.formula, s.formula, dat, ID = dat$patient)
long.dat <- l[[1]]
surv.dat <- l[[2]]
```

Head of longitudinal dataset and survival dataset
```{r}
head(long.dat)
head(surv.dat)
```

### Summary statistics for AIDS dataset

```{r}

```

### Model fit for AIDS dataset - MCEM algorithm

```{r}

```

### Model fit for AIDS dataset - Bayesian method

```{r}

```

### RSF for AIDS dataset

For RSF, since we used the CD4 count nearest to death as a covariate as well, we add this variable to our survival data.

```{r}
nearestCD4 = long.dat %>%
  select(ID, obstime, CD4, drugddI) %>%
  group_by(ID) %>%
  filter(obstime==max(obstime)) %>%
  select(ID, CD4)

surv.dat.rsf = merge(surv.dat, nearestCD4, by="ID")
s.formula.rsf <- Surv(Time, death) ~ drugddI + CD4
```

With the new formula and dataset, we fit the random survival forest.

```{r}
fit = jm.rsf(surv.dat.rsf, s.formula.rsf)

fit$predicted ## predicted value for each patient
```

```{r}
fit$survival ## In-bag survival function
```

```{r}
plot.survival(fit) ## generates Survival plots
```


### Cross-validation: AUC and BS for AIDS dataset (MCEM, Bayesian)

```{r}

```

### Cross-validation: AUC and BS for AIDS dataset (RSF)

```{r}
cv.dat <- create_cv(long.dat, surv.dat)
surv.train2 = left_join(cv.dat[[2]]$surv.train, nearestCD4, by="ID")
surv.test2 = left_join(cv.dat[[2]]$surv.test, nearestCD4, by="ID")

fit = jm.rsf(surv.train2, s.formula2)
res = predict(fit, newdata = surv.test2)
res$err.rate[!is.na(res$err.rate)] ## prediction error rate

## AUC
## BS
```

### Simulation: evaluate correctness of MCEM and Bayesian methods

```{r}

```

### Simulation: estimated AUC, true AUC and estimated BS (MCEM and Bayesian)

```{r}

```

### Simulation: estimated AUC, true AUC and estimated BS (RSF)

```{r}
```



# Discussion

# Future direction

# References 

Li, Kan, and Sheng Luo. “Dynamic Predictions in Bayesian Functional Joint Models for Longitudinal and Time-to-Event Data: An Application to Alzheimer’s Disease.” Statistical Methods in Medical Research 28, no. 2 (2017): 327–42.

https://doi.org/10.1177/0962280217722177

http://www.drizopoulos.com/vignettes/multivariate\%20joint\%20models

http://www.drizopoulos.com/vignettes/dynamic_predictions

