---
title: "Stan Tutorial"
author: "zouhaotian"
date: "3/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to Stan

Stan is a programming language that deals with full Bayesian Inference using NUTS or HMC sampling, and approximate Bayesian sampling using Variational Inference (VI), and also penalized maximum likelihood estimation with optimization.

View official website of Stan here: https://mc-stan.org/

First, let's install and load rstan:
```{r}
## install.packages('rstan')
## setwd("Stan_Tutorial/")
library(rstan)
library(nlme)
library(survival)
```

You should be able to load rstan version 2.19.3.

Some useful options:
```{r}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

The first command sets the cores to maximum cores that your PC owns. This will enable parallel sampling for multiple chains for Stan, and dramatically increase the speed.

The second command avoids re-compiling the same Stan code. Stan first translates your stan file to cpp file, and sample using this cpp file. Once the cpp file is created (in .rds suffix) and no changes are made, then this option will allow you to avoid re-compiling the stan file.

## Example 1: simple linear regression

First, we simulate 200 observations, using:

$$
\begin{aligned}
Y_i =& \beta_0 + \beta_1*X_i + \epsilon_i \\
=& 1 + 2*N(0, 1) + N(0, 1)
\end{aligned}
$$

```{r}
set.seed(1)
n <- 200
x <- rnorm(n)
y <- 1 + 2*x + rnorm(n)
lm(y ~ x)
```

Then, we write the stan code: File -> New File -> Stan File. Copy and paste the following code:

```{r, eval=FALSE}
data{
  int n;
  real x[n];
  real y[n];
}
parameters{
  real beta[2];
  real <lower=0> sigma_e;
}
transformed parameters{
  real mu[n];
  for (i in 1:n){
    mu[i] = beta[1] + beta[2]*x[i];
  }
}
model{
  beta ~ normal(0, 10);
  sigma_e ~ inv_gamma(0.1, 0.1);
  
  y ~ normal(mu, sigma_e);
}

```

Next, we compile and run the stan code:
```{r}
md <- stan_model('ex1.stan')

## Specify stan data, parameters, and initial values for parameters
stanDat <- list(n = n, x = x, y = y)
pars <- c('beta', 'sigma_e')
beta_hat <- unname(lm(y ~ x)$coefficients)
sigma_hat <- sigma(lm(y ~ x))
inits <- list(beta = beta_hat, sigma_e = sigma_hat)
inits <- list(c1 = inits, c2 = inits, c3 = inits, c4 = inits)

## sampling from stan
fitStan <- sampling(md, data = stanDat, pars = pars, init = inits, chains = 4, seed = 1)
fitStan
summary(fitStan)$summary

## Extract the samples
samples <- fitStan@sim$samples
length(samples)
length(samples[[1]]$`beta[1]`)

```

## Example 2: linear mixed effects model

We simulate longitudinal data:

$$
\begin{aligned}
Y_{ij} =& \beta_0 + \beta_1*t_{ij} + u_i + \epsilon_{ij}\\
=& 4 + 2*t_{ij} + N(0, 2) + N(0, 1)
\end{aligned}
$$

```{r}
set.seed(1)
n <- 200
obstime <- c(0, 1, 2, 3, 4)
n.time <- length(obstime)
time <- rep(obstime, n)
ID <- rep(1:n, each = n.time)

sigma_u <- sqrt(2)
sigma_e <- 1

u <- rnorm(n, 0, sigma_u)
err <- rnorm(n*n.time, 0, sigma_e)
Y <- 4 + 2*time + u[ID] + err
long.dat <- data.frame(ID = ID, time = time, Y = Y)
lme(Y ~ time, random = ~1|ID, data = long.dat)
```

Let's write the stan code:
```{r, eval=FALSE}
data{
  int n;
  int nobs;
  int ID[nobs];
  real time[nobs];
  real Y[nobs];
}
parameters{
  real beta[2];
  real <lower=0> sigma_u;
  real <lower=0> sigma_e;
  
  real u[n]; //random effects
}
transformed parameters{
  real mu[nobs];
  for (i in 1:nobs){
    mu[i] = beta[1] + beta[2]*time[i] + u[ID[i]];
  }
}
model{
  beta ~ normal(0, 10);
  sigma_u ~ inv_gamma(0.1, 0.1);
  sigma_e ~ inv_gamma(0.1, 0.1);
  
  u ~ normal(0, sigma_u);
  Y ~ normal(mu, sigma_e);
}

```

Next, we compile and run the stan code:
```{r}
md <- stan_model('ex2.stan')

## Specify stan data, parameters, and initial values for parameters
stanDat <- list(n = n, nobs = n*n.time, ID = ID, time = time, Y = Y)
pars <- c('beta', 'sigma_u', 'sigma_e')
fit <- lme(Y ~ time, random = ~1|ID)
beta_hat <- unname(fit$coefficients$fixed)
sigma_u_hat <- as.numeric(VarCorr(fit)[1, 2])
sigma_e_hat <- as.numeric(VarCorr(fit)[2, 2])
inits <- list(beta = beta_hat, sigma_u = sigma_u_hat, sigma_e = sigma_e_hat)
inits <- list(c1 = inits, c2 = inits, c3 = inits, c4 = inits)

## sampling from stan
fitStan <- sampling(md, data = stanDat, pars = pars, init = inits, chains = 4, seed = 1)
fitStan
summary(fitStan)$summary
```

## Example 3: survival outcome

We simulate 200 subjects, with hazard function as:
$$
\begin{aligned}
h_i(t) =& h_0(t)\exp(w_i \gamma)\\
=& \exp(-2)\exp(-1*Ber(0.5))
\end{aligned}
$$

Therefore, the survival function equals:
$$
\begin{aligned}
S_i(t) =& \exp(-\int_0^t h_i(t)dt)\\
=&\exp(-\exp(\log(h_0) + w_i \gamma)t)
\end{aligned}
$$

```{r}
set.seed(1)
n <- 200
logh0 <- -2
w <- rbinom(n, 1, 0.5)
gamma <- -1

s <- runif(n) ## survival probability
t <- -log(s)/exp(logh0 + w*gamma) ## failure time
c <- rep(20, n)
surv_time <- pmin(t, c)
status <- as.numeric(t<=c)
```

Next, we copy the stan code:

```{r, eval=FALSE}
data{
  int n;
  real w[n];
  real surv_time[n];
  real status[n];
}
parameters{
  real logh0;
  real gamma;
}
transformed parameters{
  real h[n];
  real S[n];
  real LL[n];
  for (i in 1:n){
    h[i] = exp(logh0 + w[i]*gamma);
    S[i] = exp(-exp(logh0 + w[i]*gamma)*surv_time[i]);
    LL[i] = log(pow(h[i], status[i])*S[i]);
  }
}
model{
  logh0 ~ normal(0, 10);
  gamma ~ normal(0, 10);
 
  target+=LL;
}

```

```{r}
md <- stan_model('ex3.stan')

## Specify stan data, parameters, and initial values for parameters
stanDat <- list(n = n, w = w, surv_time = surv_time, status = status)
pars <- c('logh0', 'gamma')
fit <- coxph(Surv(surv_time, status) ~ w, method = 'breslow')
gamma_hat <- unname(fit$coefficients)
inits <- list(logh0 = 0, gamma = gamma_hat)
inits <- list(c1 = inits, c2 = inits, c3 = inits, c4 = inits)

## sampling from stan
fitStan <- sampling(md, data = stanDat, pars = pars, init = inits, chains = 4, seed = 1)
fitStan
summary(fitStan)$summary
```